---
title: "Eco-data-science data.table workshop"
output: html_notebook

---

### shortcuts

Run Chunk : *Cmd+Shift+Enter*

Insert Chunk : *Cmd+Option+I*

Preview HTML : *Cmd+Shift+K*

# Introduction

## About the data

The datasets we will be using in this interactive session are taken from the Environmental Protection Agency's (EPA) Air Markets Program Data (AMPD) program. The first dataset (the ``epa_ampd_hourly_2019_selected.csv`` file) is a dataset of hourly emissions, electriticity generation, and fuel consumption for coal and natural gas across the country. Because the full dataset downloaded from the AMPD site is extremely large, we have filtered out the dataset to only have generators in the states of X, Y, and Z. 

The second dataset (the ``facility_01-27-2021_224024745.csv`` file) is a list of generators and various attributes, such as their fuel type, generator type, and location.

# Packages

## Install packages

Along with the ``data.table`` package, we will also be using the ``here`` package for this notebook. The ``here`` package basically helps us [write something here]. 
We also briefly use the ``janitor`` package to clean up column names.
```{r}
# install.packages("data.table")
# install.packages("here")
# install.packages("janitor")
```

## Load packages

Load the ``data.table`` and ``janitor`` packages. We don't need to load the ``here`` package because we can explicitly call on the package when using the here function (``here::here()``).
```{r}
library(data.table)
library(janitor)
```

# Read in data

## File names
Here, we are just explicitly setting objects/variables for the data files, so we can easily use them later.
```{r}
hourly_data     = 'epa_ampd_hourly_2019_selected.csv'
facility_data   = 'facility_01-27-2021_224024745.csv'
```

## Read in CSV files

Use the ``fread`` function to read in the hourly data. 
```{r}
dt = fread(here::here("data", hourly_data))
```

View first few rows of the hourly data:
```{r}
head(dt)
```
We can see that the column names are not ideal (spaces and parantheses are not easy to work with). Thus, let's use the ``janitor`` package to clean up the column names:
```{r}
dt = clean_names(dt)
head(dt)
```
Nice! Now the column names are all lower case, the spaces have been replaced with underscores, and the parantheses and slashes are gone. There are some problematic column names (for example, mmBtu has been changed to mm_btu, but we will fix those columns later).

# Basics

## Setting keys

Setting keys for the datatable can help speed up processes later, particularly for columns/keys that you frequently use (e.g., if you keep summarizing across the same groups). So, let's set the keys of ``dt`` as the orispl_code, unitid, op_date, and op_hour columns:

```{r}
setkey(dt, orispl_code, unitid, op_date, op_hour)
```

## Filter rows

In this section, we will go over how to filter the datatable's rows in a multitude of ways.

### Filter for rows for generators in California (CA)
```{r}
dt[state=="CA"]
```
### Filter for rows that occurred after july 1, 2019
```{r}
dt[op_date > "07-01-2019"]
```
### Filter for rows for facilities with the word "Energy" in its name
```{r}
dt[facility_name %like% "Energy"]
```
### Filter for rows that are within California and New Jersey
```{r}
dt[state %chin% c("CA", "NJ")]
```

## Select columns

### Select columns by name
```{r}
dt[, .(orispl_code, unitid, op_date, op_hour, op_time, gload_mw)]
```
### Select columns by column number/index
```{r}
dt[, c(3:8)]
```
# Modifying the datatable

## Rename columns

### Renaming single columns
```{r}
setnames(dt, "heat_input_mm_btu", "heat_input_mmbtu")
setnames(dt, "so2_rate_lbs_mm_btu", "so2_rate_lbs_mmbtu")
setnames(dt, "nox_rate_lbs_mm_btu", "nox_rate_lbs_mmbtu")
setnames(dt, "co2_rate_tons_mm_btu", "co2_rate_tons_mmbtu")
```

## Reorder rows

### Temporarily sorting
```{r}
dt[order(-gload_mw)]
```
### Actually sorting the data table

Order ``dt`` by the orispl_code column:
```{r}
setorder(dt, orispl_code, na.last = TRUE)
head(dt)
```
We can also order the datable by multiple columns. For example, let's say we want ``dt`` to be ordered by the following columns (in this order): state, orispl_code, unitid, op_date, and op_hour. 
```{r}
setorder(dt, state, orispl_code, unitid, op_date, op_hour)
head(dt)
```

## Summarize/aggregate

### Summarize by group(s)

Let's find the total amount of CO2 emitted by state and by date:
```{r}
dt[, .(total_co2_tons = sum(co2_mass_tons, na.rm = T)), by = .(op_date)]
```

### summarize several columns

``.SD`` can be used to summarize multiple columns
```{r}
dt[, lapply(.SD, sum, na.rm = T), .SDcols = c("co2_mass_tons", "so2_mass_lbs", "nox_mass_lbs")]
```

### summarize several columns by group(s)

We can also use ``.SD`` by group:
```{r}
dt[, lapply(.SD, sum, na.rm = T), .SDcols = c("co2_mass_tons", "so2_mass_lbs", "nox_mass_lbs"), by = .(state)]
```

### summarize multiple functions on several columns by group(s)

The ``.N`` special character can be used to count the number of observations. Not to be confused with the ``uniqueN`` function, which counts the number of *unique* observations. Let's find the number of observations, the number of power plants, and the total generation by state:
```{r}
dt[, .(number_of_observations = .N,
       number_of_plants = uniqueN(orispl_code),
       total_generation = sum(gload_mw, na.rm = T)), by = .(state)]
```

## add columns

Adding a new column that's just one same value:
```{r}
dt[, new_column := "hi"]
dt
```

You can remove a column by setting it to NULL:
```{r}
dt[, new_column := NULL]
dt
```

Change a column without creating a new one:
```{r}
dt[, gload_mw := as.numeric(gload_mw)]
```

## Replace values

Replace infinite and NA values with zeroes
```{r}
dt[is.infinite(gload_mw), gload_mw := 0]
dt[is.na(gload_mw), gload_mw := 0]
dt
```

## add columns by groups

Let's say we want to add a column that's the max generation for each generator:
```{r}
dt[, max_gen := max(gload_mw, na.rm = T), by = .(orispl_code, unitid)]
```

```{r}
head(dt)
```

### adding multiple columns
functional form vs the other form (LHS:=RHS)


## reshape between long and wide forms

### melt (wide to long)

### dcast (long to wide)

## setkey?

# joining data

## merge

## bind

# Beyond...

How much electricity generation, NOx emissions, SO2 emissions, etc are coming from coal vs natural gas plants?