---
title: "Introduction to data.table"
author: "Meas Meng and Tracey Mangin"
date: "1/28/2021"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: false
      countIncrementalSlides: false
      ratio: "16:9"
---

class: inverse, center, middle
name: background

# Who we are

<hr/>

---

# maybe we can introduce ourselves here?

---

class: inverse, center, middle
name: background

# Intro to data.table

<hr/>

---

# Why data.table?

data.table provides an enhanced, high-performance version of base R’s data.frame with syntax and feature enhancements for ease of use, convenience and programming speed.

--

Some pros of data.table:
--

* concise and consistent syntax: fast to type, fast to read

--
* fast speed

--
* memory efficient

---

# Syntax intro

Go over DT[i, j, by]

Go over the walrus :=

list() vs .()

---

class: inverse, center, middle
name: background

# EXTRA SLIDES FROM DEFAULT SHINY PRESO

<hr/>

---

# R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

---

# Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

---

# Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

---

# Slide with Plot

```{r pressure}
plot(pressure)
```

