---
title: "Introduction to data.table"
author: "Meas Meng and Tracey Mangin"
date: "1/28/2021"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: false
      countIncrementalSlides: false
      ratio: "16:9"
---

class: inverse, center, middle
name: background

# Who we are

<hr/>

---

# maybe we can introduce ourselves here?

---

# Motivation

--

MEAS: WHY IS THIS NOT WORKING

--

Tracy Manging: mind blown after running Meas's code

---

# useful resources

We used the following resources to build this preso:

---

class: inverse, center, middle
name: background

# Intro to data.table

<hr/>

---

# What is data.table?

data.table provides an enhanced, high-performance version of base R’s data.frame with syntax and feature enhancements for ease of use, convenience and programming speed.

--

## Why use data.table?
--

* concise and consistent syntax: fast to type, fast to read

--
* fast speed

--
* memory efficient

---

# What is data.table?

data.table provides an enhanced, high-performance version of base R’s data.frame with syntax and feature enhancements for ease of use, convenience and programming speed.

## Why use data.table?

* **concise and consistent syntax: fast to type, fast to read**
* fast speed
* memory efficient

---

# Syntax intro

Add details and examples for:

Go over DT[i, j, by]
(version of grant's slide 14 with tidyverse function equivalents)
(add slides on examples)

i details: filtering
helpers (%like%, %between%, %chin%)

j details: selecting
list() vs .()

by details: grouping

Go over the walrus :=
(LHS:RHS vs functional form)
(:= NULL (remove column)

modifying in place
add examples of when to expect that something is modified in place 

---

# Special symbols

.SD: **S**ubset of **D**ata
include example using less efficient version and .SD version
include tidyverse example https://atrebas.github.io/post/2019-03-03-datatable-dplyr/#summarise-data

.N: number of observations
(tidyverse equivalent n())

---

# Chaining operations

Comparisons of tidyverse vs data.table syntax

---

# What is data.table?

data.table provides an enhanced, high-performance version of base R’s data.frame with syntax and feature enhancements for ease of use, convenience and programming speed.

## Why use data.table?

* concise and consistent syntax: fast to type, fast to read
* fast speed
* memory efficient

---

# What is data.table?

data.table provides an enhanced, high-performance version of base R’s data.frame with syntax and feature enhancements for ease of use, convenience and programming speed.

## Why use data.table?

* concise and consistent syntax: fast to type, fast to read
* **fast speed**
* memory efficient

---

# fast speed stuff

fread vs read_csv

compare speed of chained operations

keys

benchmark again (dplyr vs dt vs dt with keys)

merge

write.csv vs fwrite

---

# random cool things

---

# random not so cool things

---

# final thoughts

---

class: inverse, center, middle
name: background

# EXTRA SLIDES FROM DEFAULT SHINY PRESO

<hr/>

---

# R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

---

# Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

---

# Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

---

# Slide with Plot

```{r pressure}
plot(pressure)
```

